Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-10-27 WSSSPE3/report/wg_appendix_software_credit.tex
Finished draft of credit appendix

On the second day of WSSSPE3, the working group decided to put aside the taxonomy of contributions and focus on software citations to ensure developers receive credit (regardless of contribution).
Eventually, once standardizing software citations, the goal would be to return to establishing different roles\slash contributions for this credit.

@@

There currently is no standard/accepted mechanism for citing software or receiving credit for software (akin to publications/citations).
Software is eligible for DOI assignment, but DOI metadata fields are not well tuned or standardized for software (vs. publications).
Some software providers apply for DOIs but it is still not widely adopted.
Also, there is no mechanism to cite software dependencies within software.

Obstacle: indexers (e.g., Scopus, Web of Science, Google Scholar) don't currently support software/data document types or DataCite DOIs.

Obstacle: no standard mechanism for software to cite other software.

Obstacle: for tenure, there is no standard for software products to be included in policy,
sometimes not even within one university: hard to change globally.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Key Next Steps}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Plan for Future Organization}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The WSSSPE breakout group plans to join efforts related to citing software with the FORCE11 Software Citation Working Group (FORCE11-SCWG)\footnote{FORCE11 Software Citation Working Group, \url{https://www.force11.org/group/software-citation-working-group}}; Kyle Niemeyer formally requested the merging of these groups following the meeting.
However, some future plans of the WSSSPE group fall outside the scope of FORCE11-SCWG, which covers software citation practices.
These activities include working with indexers such as Web of Science and Scopus to index software citations archived on, e.g., Zenodo\footnote{Zenodo, \url{https://zenodo.org}}, and pursuing the development of an open indexing service; such plans will be pursued either separately or through the formation of follow-on FORCE11 working groups.

>>>>

On the second day of WSSSPE3, the working group decided to put aside the taxonomy of contributions and focus on software citations to ensure developers receive credit (regardless of contribution).
Eventually, once standardizing software citations, the goal would be to return to establishing different roles\slash contributions for this credit.
Following this decision, the group identified two necessary actions to move forward:
\begin{enumerate}
    \item standardizing a citation file or some other form of metadata associated with software, and
    \item standardizing the way to cite software (used directly) in papers.
\end{enumerate}
For both of these actions, a number of ongoing efforts were identified and discussed.

\subsubsection{Software Citation Metadata}

At a minimum, the metadata needed for software citation (aside from the DOI for the software itself) includes:
\begin{itemize}
    \item Contributors, in the form of name and ORCID,
    \item Software dependencies, in the form of DOI, and
    \item Other people and artifacts that would be cited or acknowledged in a paper.
\end{itemize}
This information would then be contained in a citation file as part of 
the, e.g., GitHub repository.
The group also discussed similar efforts such as Codemeta\footnote{Codemeta: \url{https://github.com/codemeta/codemeta}}, an attempt to codify minimal 
metadata schemes in JSON and XML for scientific software and code, and 
implementing transitive credit via JSON-LD~\cite{wssspe2_katz}.
Some questions arose about how this information would be stored for 
closed-source software.

As one mechanism for constructing accurate contributor lists from existing 
project contributors, the group discussed associating GitHub accounts---as 
well as accounts on Bitbucket, CodePlex, and other repositories for 
open-source scientific software---accounts with ORCID.
However, a (quick) response from GitHub (via Arfon Smith) indicated that 
this might not be possible in the near future: ``GitHub doesn't have any 
plans to allow ORCID accounts to be associated with GitHub user accounts.''

\subsubsection{Citing Software in Publications}

Although far from a standard practice, examples of citing software in 
publications can be found in various scientific communities---notably, 
representative samples can be found in astronomy\footnote{Astronomy software citation examples: \url{https://docs.google.com/document/d/1q9ULl7alA3veL7Qwg7jGteRWeJwlrkvRHSXjvt-rTs0}} and biology~\cite{Howison2015}.
The group recommended collecting similar examples from other communities, then 
developing a software citation principles document in concert with the 
FORCE11 Software Citation Working Group (see \S\ref{SC:plan} for more details) 
following the model of the FORCE11 Data Citation Principles 
document~\cite{DataCitation2014}.

The group also discussioned briefly whether software used directly in 
a publication---whether to perform simulation, analysis, or as a dependency 
for new software developed---should be distinguished from other references 
due to the dependence of the study on these research artifacts.
Suggestions included a separate list of citations (with DOIs) for software 
and other research objects that serve this sort of ``vital'' role.
Similar recommendations were made by the credit breakout group at 
WSSSPE2~\cite{WSSSPE2}.

Finally, although a discrete task from the software citation, significant 
discussion focused on ensuring software citations are indexed in the same 
manner as publications---allowing the construction of a corresponding 
software citation network.
Currently, software releases can receive citable DOIs via 
Zenodo\footnote{Zenodo, \url{https://zenodo.org}} and 
FigShare\footnote{FigShare: \url{https://figshare.com}}; however, these 
citations are not processed by indexers such as Web of Science, Scopus, 
or Google Scholar.
Thus, either in parallel or following the primary task, the group will 
need to reach out to these organizations 
Initial conversations with Elsevier\slash Scopus (via Michael Taylor) 
during WSSSPE3 clarified that Scopus is not yet DataCite DOI aware, and 
also does not yet have an internal identifier for software or data (but 
needs\slash plans to add this support).
According to Taylor, they prefer a ``software article'' with the usual 
article metadata (e.g., authors, citations), and mentioned Zenodo as an 
example---this proposal seemed to align with group discussions.
Taylor also mentioned another benefit of the software and associated DOI 
on GitHub: in addition to a citation, could obtain statistics on 
usage/downloads/forks.

@@

Modern scientific research relies considerably on software, and ensuring 
that developers of such scientific software receive credit for their efforts 
will encourage additional creation and maintenance.
There currently is no standard mechanism for citing software or receiving credit for software (akin to publications/citations).
Software is eligible for DOI assignment, but DOI metadata fields are not well tuned or standardized for software (vs. publications).
Some software providers apply for DOIs but it is still not widely adopted.
Also, there is no mechanism to cite software dependencies within software.

Major obstacles include the fact that indexers (e.g., Scopus, Web of Science, 
Google Scholar) do not currently support software/data document types or 
DataCite DOIs.
Therefore, even with univeral association of scientific software with DOIs 
and standardizing practices for citing software in publications

Although the working group discussions at WSSSPE3 did not focus much on the 
topic of tenure, the group recognized that there is no standard policy---even 
within a single university---for software products to be included in 
promotion and tenure dossiers.
Thus, it may be difficult to encourage valuing software contributions across 
the United States or United Kingdom, and globally; furthermore, stakeholders 
are typically not tenured and thus may not be influential enough.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Key Next Steps}
\label{SC:next-steps}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Plan for Future Organization}
\label{SC:plan}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The WSSSPE breakout group plans to join efforts related to citing software with the FORCE11 Software Citation Working Group (FORCE11-SCWG)\footnote{FORCE11 Software Citation Working Group, \url{https://www.force11.org/group/software-citation-working-group}}; Kyle Niemeyer formally requested the merging of these groups following the meeting.
However, some future plans of the WSSSPE group fall outside the scope of FORCE11-SCWG, which covers software citation practices.
These activities include working with indexers such as Web of Science and Scopus to index software citations archived on, e.g., Zenodo or FigShare, and pursuing the development of an open indexing service; such plans will be pursued either separately or through the formation of follow-on FORCE11 working groups.
